SULI Report!

### 2015-08-18 doc/suli-report-swe.tex
Added the first draft of the Methods section in the SULI report. #81

In this paper, I aim to do the former by applying a simple dark matter halo model to the Millennium Simulation and predicting the weakly lensed ellipticities of generated background sources. A brief introduction into the theory of measuring galaxy ellipticities, the effects of strong and weak gravitational lensing, and dark matter halos is discussed in Section II. The implementation of a halo model on the Millennium Simulation is described in Section III, and the results of the model on (x \# of galaxies on a field size of z deg$^2$) as well as a comparison of the predicted lensed ellipticities to the true ellipticities is given in Section IV.  Section V discusses limitations of the Pangloss framework as well as potential physical and computational improvements that can be made for upcoming work before concluding remarks in Section VI.

\section{Theory}

@@

\section{Methods}

\begin{itemize}
\item Assumptions
    \begin{itemize}
    \item A catalog with galaxy ellipticities
    \item Currently use actual dark matter halo mass for an idea of best case scenario - can run same analysis with masses sampled from SMHR (reference)
    \item PSF already accounted for in catalog and error low enough for weak lensing analysis
    \item Use actual redshift for best case scenario (but can easily be replaced by photometric redshift sampling)
    \end{itemize}
\item Millennium Simulation
    \begin{itemize}
    \item Instead of blindly using \texttt{Pangloss} to make dark matter mass maps using sky survey data, we instead test how accurately and precisely we can predict weakly-lensed shear and convergence 
    \item Mention which M.S. we used
    \item S. Hilbert`s ray traced M.S. (reference)
        \begin{itemize}
        \item Can use the ray-traced convergence and shear to calculate the ``true'' lensed ellipticties and compare to \texttt{Pangloss}
        \end{itemize}
    \end{itemize}
\item Background galaxies
    \begin{itemize}
    \item random orientation
    \item ellipticity range (get numbers!)
    \item Accounting for shape noise
    \end{itemize}
\item Lightcones
    \begin{itemize}
    \item Instead of using all foreground mass in a field to calculate the lensing of a single background object, we create a lightcone centered at the source with a chosen radius $R$
    \item Drilling
        \begin{itemize}
        \item redshift grid to speed up calcuations
        \item Use the physical distance to line of sight in lensing calculations
            \begin{itemize}
            \item Redshift-to-distance relation (equation and reference)
            \end{itemize}
        \item Ranking galaxy `relevance'
            \begin{itemize}
            \item Want to only include the most important galaxies for lensing to speed up computation, as most will have negligable mass or be too far away
            \item Allows quicker lensing calculations, means we can make $R$ bigger
            \item Show `relevance` metric ($M/R^3$ but unitless)
            \item Include plot of predicted relevance vs kappa contribution??
            \end{itemize}
        \end{itemize}
    \item Lensing by Halos
        \begin{itemize}
        \item Treat each foreground halo as a separate lensing event
        \item Total lensing is the sum of each event
        \item Lookup table??
        \end{itemize}
    \item Include cartoon
    \end{itemize}
\item Checking the halo model: Use of correlation functions as summary statistics for ellipticity predictions
    \begin{itemize}
        \item Introduce the concept of correlation functions in general
            \begin{itemize}
            \item Definition
            \item Autocorrelation vs cross-correlation
            \item Use of correlation functions in cosmology
            \item Shear-shear and galaxy-galaxy correlation functions
                \begin{itemize}
                \item Reference appendix for Visualizing Correlation Functions
                \end{itemize}
            \end{itemize}
        \item Use of Jarvis`s \texttt{TreeCorr} code to calculate cf`s
        \begin{itemize}
            \item Reference appendix for Jarvis proof
        \end{itemize}
    \end{itemize}
\item Show one map with ray-traced features and lensed ellipticities vs another map with \texttt{Pangloss} features and predicted ellipticities?
\end{itemize}

@@

\begin{itemize}
\item Issues with the \texttt{Pangloss} framework
    \begin{itemize}
    \item Systematic shear power under-prediction in cf`s
    \item Missing large-scale dark matter structure
    \end{itemize}
    \item Possibly missing features in model
        \begin{itemize}
        \item voids
        \item filaments
        \item stellar mass
        \end{itemize}
    \item Speed
\item Moving Foward
    \begin{itemize}
    \item Scaling up
        \begin{itemize}
        \item CPU and RAM numbers for lightcone parallelization from discussion with Debbie Bard
        \end{itemize}
    \item other stuff - too tired right now
    \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item P.J. Marshall Thesis?
\item Collett et al.
\item S. Hilbert (ray-traced)
\item Millennium Simulation
\item Schneider
\item Jarvis \texttt{TreeCorr}
\item NFW profile
\item BMO truncation
\item Planck
\item galaxy curves
\item BAOs
\item LCDM
\item GR
\item wl distortion percent
\end{itemize}

@@

%% Intro:
\begin{itemize}
\item Broad intro into cosmology and dark matter, set problem in context
    \begin{itemize}
    \item (Very) short introduction to history of dark matter, Zwicky, rotation curves, etc.
    \end{itemize}
\begin{itemize}
\item Introduction to gravitational lensing
    \begin{itemize}
    \item Convergence and Shear / Strong and Weak lensing
        \begin{itemize}
        \item Dramatic effect of strong lensing, how it is used for constraining cos. parameters
        \item Only a few hundred known, not good for making mass maps
        \item All galaxies are weakly lensed, good for our purposes
        \end{itemize}
    \end{itemize}
\end{itemize}
\item What this work does and aims for
\begin{itemize}
    \item We aim to implement a simple dark matter halo model to predict weak-lensing effects by foreground mass - and to make the prediction `quickly' and/or with large parallelization
    \item Quick summary of what we do in paper
        \begin{itemize}
        \item Use halo model to make approximation of dark matter mass distribution in Millennium Simulation
        \item Use shear-shear and galaxy-galaxy correlation functions as summary statistics
        \item Compare statistics to ray-traced data
        \item Discuss physial and computational shortcomings with current methodology
        \end{itemize}
    \end{itemize}
\end{itemize}

>>>>

In this paper, I aim to do the former by applying a simple dark matter halo model to the Millennium Simulation and predicting the weakly lensed ellipticities of generated background sources. A brief introduction into the theory of measuring galaxy ellipticities, the effects of strong and weak gravitational lensing, and dark matter halos is discussed in Section II. The implementation of a halo model on the Millennium Simulation is described in Section III, and the results of the model on (x \# of galaxies on a field size of x \# deg$^2$) as well as a comparison of the predicted lensed ellipticities to the true ellipticities is given in Section IV.  Section V discusses limitations of the used Pangloss framework as well as potential physical and computational improvements that can be made for upcoming work before concluding remarks in Section VI.

\section{Theoretical Background}

@@

\section{The Pangloss Framework}

To constrain the mass of foreground dark matter using weak gravitational lensing, first a model of the relationship between foreground galaxies and the foreground dark matter must be established and robustly tested to see if, statistically, it makes the same lensing predictions of background sources as the true undering dark matter structure. To do this, I built upon the publically available \texttt{Pangloss} framework used in Collett et al. (cite) to reconstruct all the mass along the line of sight of each background galaxy using dark matter halos. The lensing contribution of each halo is calculated, and the total lensing of the background galaxy is the sum of each halo contribution. This process is detailed in the following sections.

\subsection*{Assumptions}
While \texttt{Pangloss} may be used more generally, the present analysis makes some additional strong assumptions to simplify the problem for a first attempt at making weak lensing predictions.

\begin{enumerate}
\item The dark matter mass distribution can be approximated by spherically symmetric BMO halo profiles attached to each galaxy.
\item The stellar mass of the foreground galaxy is negligible for lensing calculations.
\item The mass of the dark matter halo of each foreground galaxy is known.
\item A spectroscopic redshift of each foreground galaxy is known.
\end{enumerate}

Testing the first assumption is the main goal of this paper. The second assumption is reasonable as it is estimated that dark matter halos are 1-2 orders of magnitude more massive than the host galaxy (cite). Clearly the third assumption will never be true for any observational data. However, this allows for a best-case scenario estimate of how well the \texttt{Pangloss} framework \textit{could} predict weak lensing effects given all possible information. This assumption can be relaxed by sampling a dark matter halo mass from an assumed stellar mass - halo mass relation (cite). The fourth assumpition is also unrealistic as most galaxies in sky surveys only have a less-reliable photometric redshift due to time constraints, but this again allows for a best-case estimate. This assumption could be relaxed by adding random photometric redshift noise and repeating the upcoming analysis on many realizations of the photometric redshifts.

\subsection*{The Millennium Simulation}
\texttt{Pangloss} cannot be used to make dark matter mass maps using existing galaxy catalogs until it is tested on a simulated universe with known dark matter structure to determine how accurately and precisely it predicts the lensing of background sources. For this purpose, \texttt{Pangloss} was tested on galaxy catalogs from the Millennium Simulation, an N-body simulation consisting of over 10 billion dark matter `particles' each representing a billion solar masses and populated with about 20 million galaxies (cite 2005). The simulation uses cosmological parameters from WMAP 1\textsuperscript{st}-year data analysis and contains baryonic and dark matter structure believed to be consistent with our own universe. From the work of Hilbert et al. (cite), high resolution maps of the shear and convergence from the perspective of a single reference point calculated by ray-tracing are publically available. From these maps, the actual lensing of background galaxies when traveling through the foreground mass of the Millennium Simulation towards the reference point can be calculated using Equation (??).

\subsection*{Generating Background Galaxies}
With a catalog of foreground galaxies chosen, a set of background galaxies with density 10 per square arcminute was generated. The intrinsic orientation of each galaxy was sampled from a uniform distribution as, without lensing, there should be no preferred orientation due to the assumption of an isotropic universe. The magnitude of the galaxy ellipticities was sampled from a normal distribution with a standard deviation of 0.2, but any ellipticities with magnitude greater than one were re-sampled. Random ellipticity noise was sampled from a normal distribution with a standard deviation of 0.1 and added to the intrinsic ellipticities. Finally, each ellipticity was multiplied by 0.9 to account for systematic shape noise of 10\%.

\subsection*{Creating Lightcones}
While ideally all foreground mass in a field would be considered when predicting the weak lensing of a background galaxy, it is computationally prohibitive to do so. Instead, all foreground halos contained within a `lightcone' centered along the line of sight to the source and extending out to a chosen lightcone radius $R$ were considered when calculating lensing contributions for the background source. A cartoon model of this process is shown in Figure (???). Unless otherwise specified, experiments in this paper used a radius of 2 arcminutes.

%************************
(Pangloss Cartoon Figure)
%************************

To calculate the convergence and shear contribution of each halo, the physical distance from the halo to the line of sight was needed. To increase the speed of distance calculations, the foreground halo redshifts were first snapped to a grid of 100 equally-sized redshift bins and then converted to physical distance using (still trying to decipher the grid.py/distance.py code. Ask Phil!).

%************************
(Redshift-to-distance relation equation?)
%************************

Using the physical separation distance and halo mass, the shear and convergence of a single foreground halo is calculated using methods described in Wright and Brainerd (cite). The total convergence and shear at the center of the lightcone is simply the sum of the convergence and shear contributions of each halo contained in the lightcone. The \texttt{Pangloss}-predicted lensed background ellipticity is then calculated using Equation (??).

\subsection*{Checking the Halo Model}

Instead of comparing the ray-traced and \texttt{Pangloss}-predicted lensed ellipticities for individual galaxies, the lensing is characterized globally with correlation functions. The ellipticity-ellipticity correlation function measures how correlated the ellipticities of pairs of galaxies are as a function of separation distance, while the galaxy-galaxy correlation function measures the correlation of lensed ellipticities around foreground halos as a function of separation distance. For readers that are unfamiliar with correlation functions in the context of cosmology or want a visual aid, see Appendix A. Both correlation functions are used in this work to measure how well the \texttt{Pangloss} framework models weak lensing by dark matter structures using the publically available \texttt{TreeCorr}\footnote{https://github.com/rmjarvis/TreeCorr} module written by Mike Jarvis. Note that the correlation function definition used in \texttt{TreeCorr} is slightly different than that used in most of the literature; for a derivation of the connection between Jarvis's definition and the more common Schneider definition (cite), see Appendix B.

@@

\begin{itemize}
\item Issues with the \texttt{Pangloss} framework
    \begin{itemize}
    \item Systematic shear power under-prediction in cf`s
    \item Missing large-scale dark matter structure
    \end{itemize}
    \item Possibly missing features in model
        \begin{itemize}
        \item voids
        \item filaments
        \item stellar mass
        \end{itemize}
    \item Speed
\item Moving Foward
    \begin{itemize}
    \item Scaling up
        \begin{itemize}
        \item CPU and RAM numbers for lightcone parallelization from discussion with Debbie Bard
        \end{itemize}
    \item infer hyper-parameters from SMHR, physical distance, etc
    \item other stuff - too tired right now
    \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item P.J. Marshall Thesis?
\item Collett et al.
\item S. Hilbert (ray-traced) http://arxiv.org/abs/0809.5035
\item Millennium Simulation
\item Schneider
\item Jarvis \texttt{TreeCorr}
\item lensing calculation http://arxiv.org/pdf/astro-ph/9908213v1.pdf
\item NFW profile
\item BMO truncation
\item Planck
\item galaxy curves
\item BAOs
\item LCDM
\item GR
\item wl distortion percent
\end{itemize}

@@

%% Introduction:
\begin{itemize}
\item Broad intro into cosmology and dark matter, set problem in context
    \begin{itemize}
    \item (Very) short introduction to history of dark matter, Zwicky, rotation curves, etc.
    \end{itemize}
\begin{itemize}
\item Introduction to gravitational lensing
    \begin{itemize}
    \item Convergence and Shear / Strong and Weak lensing
        \begin{itemize}
        \item Dramatic effect of strong lensing, how it is used for constraining cos. parameters
        \item Only a few hundred known, not good for making mass maps
        \item All galaxies are weakly lensed, good for our purposes
        \end{itemize}
    \end{itemize}
\end{itemize}
\item What this work does and aims for
\begin{itemize}
    \item We aim to implement a simple dark matter halo model to predict weak-lensing effects by foreground mass - and to make the prediction `quickly' and/or with large parallelization
    \item Quick summary of what we do in paper
        \begin{itemize}
        \item Use halo model to make approximation of dark matter mass distribution in Millennium Simulation
        \item Use shear-shear and galaxy-galaxy correlation functions as summary statistics
        \item Compare statistics to ray-traced data
        \item Discuss physial and computational shortcomings with current methodology
        \end{itemize}
    \end{itemize}
\end{itemize}

%% Theoretical Background:

%% The Pangloss Framework:
\begin{itemize}
\item Assumptions
    \begin{itemize}
    \item A catalog with galaxy
    \item Currently use actual dark matter halo mass for an idea of best case scenario - can run same analysis with masses sampled from SMHR (reference)
    \item PSF already accounted for in catalog and error low enough for weak lensing analysis
    \item Use actual redshift for best case scenario (but can easily be replaced by photometric redshift sampling)
    \end{itemize}
\item Millennium Simulation
    \begin{itemize}
    \item Instead of blindly using \texttt{Pangloss} to make dark matter mass maps using sky survey data, we instead test how accurately and precisely we can predict weakly-lensed shear and convergence 
    \item Mention which M.S. we used
    \item S. Hilbert`s ray traced M.S. (reference)
        \begin{itemize}
        \item Can use the ray-traced convergence and shear to calculate the ``true'' lensed ellipticties and compare to \texttt{Pangloss}
        \end{itemize}
    \end{itemize}
\item Background galaxies
    \begin{itemize}
    \item random orientation
    \item ellipticity range (get numbers!)
    \item Accounting for shape noise
    \end{itemize}
\item Lightcones
    \begin{itemize}
    \item Instead of using all foreground mass in a field to calculate the lensing of a single background object, we create a lightcone centered at the source with a chosen radius $R$
    \item Drilling
        \begin{itemize}
        \item redshift grid to speed up calcuations
        \item Use the physical distance to line of sight in lensing calculations
            \begin{itemize}
            \item Redshift-to-distance relation (equation and reference)
            \end{itemize}
        \item Ranking galaxy `relevance'
            \begin{itemize}
            \item Want to only include the most important galaxies for lensing to speed up computation, as most will have negligable mass or be too far away
            \item Allows quicker lensing calculations, means we can make $R$ bigger
            \item Show `relevance` metric ($M/R^3$ but unitless)
            \item Include plot of predicted relevance vs kappa contribution??
            \end{itemize}
        \end{itemize}
    \item Lensing by Halos
        \begin{itemize}
        \item Treat each foreground halo as a separate lensing event
        \item Total lensing is the sum of each event
        \item Lookup table??
        \end{itemize}
    \item Include cartoon
    \end{itemize}
\item Checking the halo model: Use of correlation functions as summary statistics for ellipticity predictions
    \begin{itemize}
        \item Introduce the concept of correlation functions in general
            \begin{itemize}
            \item Definition
            \item Autocorrelation vs cross-correlation
            \item Use of correlation functions in cosmology
            \item Shear-shear and galaxy-galaxy correlation functions
                \begin{itemize}
                \item Reference appendix for Visualizing Correlation Functions
                \end{itemize}
            \end{itemize}
        \item Use of Jarvis`s \texttt{TreeCorr} code to calculate cf`s
        \begin{itemize}
            \item Reference appendix for Jarvis proof
        \end{itemize}
    \end{itemize}
\item Show one map with ray-traced features and lensed ellipticities vs another map with \texttt{Pangloss} features and predicted ellipticities?
\end{itemize}

Maybe do importance and/or lookup table after paper is submitted? Talk about in discussion?
